One-armed spirals in locally isothermal, radially structured self-gravitating discs

### 2015-01-01 paper/discussion.tex
major restructuring and rewrite in response to report

\subsection{Angular momentum density of  low-frequency $m=1$ modes}
From Eq. \ref{lin_ang_mom_cons} and assuming a time-dependence of the
form $\exp{(-\ii \sigma t)}$ with $\real{\sigma} = \omega$,  
the angular momentum density associated with linear waves is
\begin{align}
  \jlin = \frac{m\Sigma}{2}\left[\left(\omega -
      m\Omega\right)|\bm{\xi}|^2 + \ii\Omega\left(\xi_R\xi_\phi^* -
      \xi_R^*\xi_\phi\right)\right].  
\end{align}
For a low-frequency mode, $|\omega|\ll m\Omega$. Then neglecting the
term $\omega|\bm{\xi}|^2$ and setting $m=1$,
\begin{align}
  \jlin &\simeq \frac{\Sigma\Omega}{2}\left[-|\bm{\xi}|^2 + \ii\left(\xi_R\xi_\phi^* -
      \xi_R^*\xi_\phi\right)\right]\notag\\
  & = -\frac{\Sigma\Omega}{2}|\xi_R + \ii\xi_\phi|^2\notag\\
  & \leq 0.
\end{align}
Thus, low-frequency $m=1$ modes generally have negative
angular momentum. These are the modes observed in our simulations. 
If the mode loses (positive) angular momentum 
to the background, then we can expect instability. We demonstrate
below how this is possible through a forced temperature
gradient.

\subsection{Unstable interaction between low-frequency $m=1$ spirals 
  and the background disc due to an imposed temperature gradient}
The $m=1$ spiral that emerges in our simulations have low frequency
($\Omega_p < \Omega$), grows slowly ($\gamma\ll\Omega_p$) and its
spatial form is consistent local theory (\S\ref{fargo_m1}). Then to a
first approximation, we can regard it   
as a mode within the framework of local theory, satisfying the
% temperature gradient ignored in local theory
dispersion relation Eq. \ref{dispersion}. To demonstrate possible
instability, we show that the torque density acting on such a mode due
to the background temperature gradient is negative, which enforces the
mode, because it has negative angular momentum.

% Note that
% we will not be assuming a self-gravitating disc until the end of this
% section.

We begin by recalling the expression for the torque density due to a
locally isothermal equation of state,
\begin{align*}
  T_\mathrm{BG} = - \frac{m}{2}\frac{dc_s^2}{dR}\imag\left(\delta\Sigma_m\xi_R^*\right).   
\end{align*}
The Eulerian surface density perturbation is given by
\begin{align}\label{den_pert}
  \delta\Sigma_m = -\nabla\cdot\left(\Sigma\bm{\xi}\right) 
  = -\frac{1}{R}\frac{d}{dR}\left(R\Sigma \xi_R\right) - \frac{\ii m}{R}\Sigma\xi_\phi.  
\end{align}
We now invoke local theory by setting $d/dR \to \ii k$ where $k$ is
real, and assume $|kR|\gg m$ so that the second term on the right hand
side of Eq. \ref{den_pert} can be neglected. Then    
\begin{align}
  \delta\Sigma _m  \simeq -\ii k \Sigma \xi_R,
\end{align}
which gives
\begin{align}
  T_\mathrm{BG} = \frac{m}{2}\frac{dc_s^2}{dR}k\Sigma |\xi_R|^2. 
\end{align}
% \begin{align}
%   T_\mathrm{BG} &= -\frac{m}{2}\frac{dc_s^2}{dR}\imag\left[\delta\Sigma_m \left(\frac{\delta
%         v_{Rm}}{-\ii\sbar}\right)^*\right] \notag\\
%   & = \frac{m}{2}\frac{dc_s^2}{dR}\imag\left[\ii \delta\Sigma_m \left(\frac{\delta
%         v_{Rm}}{\sbar}\right)^*\right]
% \end{align}
% where we replaced the radial Lagrangian displacement by the Eulerian
% radial velocity perturbation in Eq. \ref{baroclinic_torque}. 
% The linearised momentum equations give
% \begin{align}
%   D\delta v_{Rm} =& 
%   \ii\sbar\left[c_s^2\frac{d}{dR}\left(\frac{\delta\Sigma_m}{\Sigma}\right)
%     + \frac{d}{dR}\delta\Phi_m\right] \notag\\ 
%   &- \frac{2\ii
%     m\Omega}{R}\left(c_s^2\frac{\delta\Sigma_m}{\Sigma} +
%     \delta\Phi_m\right),
% \end{align}
% where $D\equiv \kappa^2 - \sbar^2$. 
% We now invoke local theory by setting $d/dR \to \ii k$ where $k$ is
% real, and using the local solution to the Poisson equation
% \begin{align}
%   \delta\Phi_m = -\frac{2\pi G}{|k|}\delta\Sigma_m 
% \end{align}
% \citep{shu91}. Then the radial velocity perturbation becomes
% \begin{align}
%   \delta v_{Rm} =
%   \frac{1}{D}\frac{\delta\Sigma_m}{\Sigma}\left(\frac{2\pi G
%       \Sigma}{|k|} - c_s^2\right)\left(k\sbar + \frac{2\ii
%       m\Omega}{R}\right),
% \end{align}
% so that 
% \begin{align}
%   \ii\delta\Sigma_m \left(\frac{\delta v_{Rm}}{\sbar}\right)^* =
%   \frac{1}{D^*}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
%       \Sigma}{|k|} - c_s^2\right)\left(\ii k  + \frac{2
%       m\Omega}{R\sbar^*}\right). 
% \end{align}
% For a neutral mode, $\sbar$, and therefore $D$ is real. Then, after
% taking the imaginary part of the above expression, the torque
% density is 
% \begin{align}\label{tbg_explicit}
%   T_\mathrm{BG} &=
%   \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{D}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
%       \Sigma}{|k|} - c_s^2\right)\right].
% % &\simeq \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{\left(\kappa^2 -
% %         m^2\Omega^2\right)}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi
% %       G \Sigma}{|k|} - c_s^2\right)\right]. 
% \end{align}
% We now use the local dispersion relation,
% Eq. \ref{dispersion}, to make the replacement $D = 2\pi G \Sigma |k| -
% k^2c_s^2$ in the above expression. This gives
% \begin{align}\label{tbg_simple}
%   T_\mathrm{BG} =
%   \frac{m}{2}\frac{dc_s^2}{dR}\frac{1}{k}\frac{|\delta\Sigma_m|^2}{\Sigma}.  
% \end{align}  
This torque density is negative for trailing waves in discs with
a fixed temperature profile decreasing outwards ($k>0$ and $dc_s^2/dR<0$,
respectively). Note that this conclusion does not rely on the
low-frequency approximation and holds for any $m$.

However, if the linear disturbance under consideration \emph{is} a
low-frequency $m=1$ mode, then it has negative angular
momentum. If it is a trailing spiral and $dc_s^2/dR<0$, as is typical
in astrophysical discs, then $T_\mathrm{BG}<0$, and
the background disc applies a negative torque on the disturbance,
which further decreases its angular momentum. This suggests the spiral
amplitude will grow.

Using $\jlin$ and $T_\mathrm{BG}$, we can estimate the 
growth rate $\gamma$ of linear perturbations due to the background
torque as  
\begin{align}
  2\gamma \sim \frac{T_\mathrm{BG}}{\jlin},
\end{align}
where the factor of two accounts for the fact that the angular momentum
density is quadratic in the linear perturbations. Inserting the above
expressions for $\jlin$ and $T_\mathrm{BG}$ for $m=1$ gives
\begin{align}\label{theoretical_rate0}
  2\gamma \sim
  -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R|^2}{|\xi_R + \ii
    \xi_\phi|^2}. 
%  \lesssim -\frac{dc_s^2}{dR}\frac{k}{\Omega}.
\end{align}
In Appendix \ref{horizontal_displacements}, we argue $\xi_\phi \simeq
2\ii \xi_R$ for the modes of interest. Then 
%To further simplify this expression, we need to relate $\xi_R$ and
%$\xi_\phi$. 
% \begin{align}
%   2\gamma \sim
%   -\frac{dc_s^2}{dR}\frac{1}{k\Sigma^2\Omega}\frac{|\delta\Sigma_1|^2}{|\xi_R
%     + \ii\xi_\phi|^2}.
% \end{align}
% Now, using $\delta\Sigma_m/\Sigma = -\ii k \xi_R - \ii m \xi_\phi/R$
% in the local approximation, we have
% \begin{align}
%   2\gamma \sim -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R +
%     \xi_\phi/kR|^2}{|\xi_R + \ii \xi_\phi|^2}
%   \lesssim -\frac{dc_s^2}{dR}\frac{k}{\Omega}.
% \end{align}
% where we used $dc_s^2/dR <0$, $k>0$ and $|kR|\gg 1$ to obtain the
% inequality. 
%where we used $dc_s^2/dR <0$ and $k>0$ to obtain the inequality. 
for the temperature profiles $c_s^2 = c_{s0}^2 (R/R_0)^{-q}$ as
adopted in our disc models,  
\begin{align}
  2\gamma \sim q\frac{c_s^2}{R}\frac{k}{\Omega}. 
\end{align}
For perturbations in a self-gravitating disc with characteristic
wavenumber $k \sim \pi G \Sigma/c_s^2 \sim \Omega/Q c_s$, as seen in
our simulations, we have 
\begin{align}\label{theoretical_rate}
  \gamma \sim \frac{qh}{2Q}\Omega,
\end{align}
where we used $c_s/R\sim h\Omega$. For our fiducial disc models with
$q=1$, $Q=O(1)$, $h=0.05$, this gives $\gamma/\Omega \sim O(10^{-2})$,
consistent with simulation results. Specifically,
Eq. \ref{theoretical_rate} indicates the growth rate should increase
linearly with $q$ and $h$, as seen in the simulations
(Fig. \ref{fargo_varq_growth}---\ref{fargo_varh_growth}).     
%the latter assumes  k=pi\Sigma G /c_s^2 is always the most favoured

>>>>

%the latter assumes  k=pi\Sigma G /c_s^2 is always the most favoured
